---
title: "FinalProject"
output: html_document
date: "2024-12-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = FALSE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(skimr)
library(stringr)
library(ggpubr)
library(flextable)
library(readr)
library(knitr)
library(naniar)
library(plotly)
```

```{r}

games <- read.csv("/Users/williamgifford/Final Project Submission/Data/games.csv")
player_play <- read.csv("/Users/williamgifford/Final Project Submission/Data/player_play.csv")
players <- read.csv("/Users/williamgifford/Final Project Submission/Data/players.csv")
plays <- read.csv("/Users/williamgifford/Final Project Submission/Data/plays.csv")

```

# Data Dictionaries 
```{r, echo=FALSE}
# Create the data dictionary as a data frame
games_dict <- data.frame(
  Variable = c(
    "gameId", "season", "week", "gameDate", "gameTimeEastern",
    "homeTeamAbbr", "visitorTeamAbbr", "homeFinalScore", "visitorFinalScore"
  ),
  Description = c(
    "Game identifier, unique (numeric)",
    "Season of game (numeric)",
    "Week of game (numeric)",
    "Game Date (time, mm/dd/yyyy)",
    "Start time of game (time, HH:MM:SS, EST)",
    "Home team three-letter code (text)",
    "Visiting team three-letter code (text)",
    "The total amount of points scored by the home team in the game (numeric)",
    "The total amount of points scored by the visiting team in the game (numeric)"
  )
)
kable(
  games_dict,
  caption = "Table: Data Dictionary for `games.csv`"
)

```

```{r, echo=FALSE}
# Create the data dictionary as a data frame
players_dict <- data.frame(
  Variable = c(
    "nflId", "height", "weight", "birthDate", 
    "collegeName", "position", "displayName"
  ),
  Description = c(
    "Player identification number, unique across players (numeric)",
    "Player height (text)",
    "Player weight (numeric)",
    "Date of birth (YYYY-MM-DD)",
    "Player college (text)",
    "Official player position (text)",
    "Player name (text)"
  )
)

kable(
  players_dict,
  caption = "Table: Data Dictionary for `players.csv`"
)

```

```{r echo= FALSE}

# Create the data dictionary as a data frame
plays_dict <- data.frame(
  Variable = c(
    "gameId", "playId", "playDescription", "quarter", "down", "yardsToGo", 
    "possessionTeam", "defensiveTeam", "yardlineSide", "yardlineNumber", 
    "gameClock", "preSnapHomeScore", "preSnapVisitorScore", "playNullifiedByPenalty",
    "absoluteYardlineNumber", "preSnapHomeTeamWinProbability", 
    "preSnapVisitorTeamWinProbability", "expectedPoints", "offenseFormation",
    "receiverAlignment", "playClockAtSnap", "passResult", "passLength", 
    "targetX", "targetY", "playAction", "dropbackType", "dropbackDistance",
    "passLocationType", "timeToThrow", "timeInTackleBox", "timeToSack",
    "passTippedAtLine", "unblockedPressure", "qbSpike", "qbKneel", 
    "qbSneak", "rushLocationType", "penaltyYards", "prePenaltyYardsGained", 
    "yardsGained", "homeTeamWinProbabilityAdded", "visitorTeamWinProbabilityAdded",
    "expectedPointsAdded", "isDropback", "pff_runConceptPrimary", 
    "pff_runConceptSecondary", "pff_runPassOption", "pff_passCoverage", 
    "pff_manZone"
  ),
  Description = c(
    "Game identifier, unique (numeric)", "Play identifier, not unique across games (numeric)", 
    "Description of play (text)", "Game quarter (numeric)", "Down (numeric)", 
    "Distance needed for a first down (numeric)", 
    "Team abbr of team on offense with possession of ball (text)", 
    "Team abbr of team on defense (text)", 
    "3-letter team code corresponding to line-of-scrimmage (text)", 
    "Yard line at line-of-scrimmage (numeric)", "Time on clock of play (MM:SS)", 
    "Home score prior to the play (numeric)", "Visiting team score prior to the play (numeric)", 
    "Whether or not an accepted penalty on the play cancels the play outcome (text)", 
    "Distance from end zone for possession team (numeric)", 
    "The win probability of the home team before the play (numeric)", 
    "The win probability of the visiting team before the play (numeric)", 
    "Expected points on this play (numeric)", 
    "Formation used by possession team (text)", 
    "Enumerated as 0x0, 1x0, 1x1, 2x0, 2x1, 2x2, 3x0, 3x1, 3x2 (text)", 
    "What the play clock value was at time of snap (numeric)", 
    "Dropback outcome of the play (text)", 
    "The distance beyond the LOS that the ball traveled not including yards into the endzone. If thrown behind LOS, the value is negative. (numeric)", 
    "The x-coordinate of the targeted receiver when the pass arrived (numeric)", 
    "The y-coordinate of the targeted receiver when the pass arrived (numeric)", 
    "Boolean indicating whether there was play-action on the play (Boolean)", 
    "The type of drop back after the snap by the QB (text)", 
    "The distance the QB dropped back (yards) behind the center after the snap (numeric)", 
    "The location type of where the QB was at the time of throw (text)", 
    "The time (secs) elapsed between snap and pass (numeric)", 
    "The amount of time the QB spent inside the tackle box (numeric)", 
    "The time from snap to the time the QB was sacked (numeric)", 
    "Boolean indicating whether the pass was tipped at the line of scrimmage (Boolean)", 
    "Boolean indicating whether there was pressure from an unblocked player (Boolean)", 
    "Boolean indicating whether the play was a QB Spike (Boolean)", 
    "Whether or not the play was a QB Kneel (numeric)", 
    "Whether or not the play was a QB Sneak (numeric)", 
    "The direction the runner ran based on where the offensive linemen were during the play (text)", 
    "Yards gained by offense by penalty (numeric)", 
    "Net yards gained by the offense, before penalty yardage (numeric)", 
    "Net yards gained by the offense, including penalty yardage (numeric)", 
    "Win probability delta for home team (numeric)", 
    "Win probability delta for visitor team (numeric)", 
    "Expected points added on this play (numeric)", 
    "Boolean indicating whether the QB dropped back, meaning the play resulted in a pass, sack, or scramble (Boolean)", 
    "The primary run concept on the play (text)", 
    "The secondary run concept on the play (text)", 
    "Whether or not the play was a run-pass option (numeric)", 
    "The pass coverage concept employed by the defense on the play (text)", 
    "Whether the defense employed man or zone coverage on the play (text)"
  )
)

# Display the table
kable(
  plays_dict,
  caption = "Table: Data Dictionary for `plays.csv`"
)

```

```{r, echo=FALSE}
player_play_dict <- data.frame(
  Variable = c(
    "gameId", "playId", "nflId", "teamAbbr", "hadRushAttempt",
    "rushingYards", "hadDropback", "passingYards", "sackYardsOffense", 
    "hadPassReception", "receivingYards", "wasTargettedReceiver", 
    "yardageGainedAfterTheCatch", "fumbles", "fumbleLost", "fumbleOutOfBounds", 
    "assistedTackle", "forcedFumbleAsDefense", "halfSackYardsAsDefense", 
    "passDefensed", "quarterbackHit", "sackYardsAsDefense", "safetyAsDefense", 
    "soloTackle", "tackleAssist", "tackleForALoss", "tackleForALossYardage", 
    "hadInterception", "interceptionYards", "fumbleRecoveries", 
    "fumbleRecoveryYards", "wasInitialPassRusher", "penaltyNames", 
    "causedPressure", "timeToPressureAsPassRusher", "getOffAsPassRusher", 
    "inMotionAtBallSnap", "shiftSinceLineset", "motionSinceLineset", 
    "wasRunningRoute", "routeRan", "blockedPlayerNFLId1", 
    "blockedPlayerNFLId2", "blockedPlayerNFLId3", "pressureAllowedAsBlocker", 
    "timeToPressureAllowedAsBlocker", "pff_defensiveCoverageAssignment", 
    "pff_primaryDefensiveCoverageMatchupNflId", 
    "pff_secondaryDefensiveCoverageMatchupNflId"
  ),
  Description = c(
    "Game identifier, unique (numeric)",
    "Play identifier, not unique across games (numeric)",
    "Player identification number, unique across players (numeric)",
    "The team abbreviation for the team the player plays for (text)",
    "Whether or not the player had a rushing attempt on this play (numeric)",
    "The rush yards accrued by the player on this play (numeric)",
    "Whether or not the player dropped back on this play (numeric)",
    "The pass yards accrued by the player on this play (numeric)",
    "The yards lost by the player via a sack on this play (numeric)",
    "Whether or not the player caught a pass on this play (numeric)",
    "The receiving yards accrued by the player on this play (numeric)",
    "Whether or not the player was the intended receiver on this play (numeric)",
    "The yards gained after the catch was made by the player on this play (numeric)",
    "The number of fumbles by the player on this play (numeric)",
    "Whether or not the player lost a fumble to the opposing team on this play (numeric)",
    "Whether or not the player fumbled the ball out of bounds on this play (numeric)",
    "Whether or not the player required an assist to make a tackle on this play (numeric)",
    "Whether or not the player forced a fumble by the opposing team on this play (numeric)",
    "The yards conceded by the offense because of a half-sack by the player on this play (numeric)",
    "Whether or not a passing play was stopped by the player on this play (numeric)",
    "Whether or not the player recorded a QB hit on this play (numeric)",
    "The yards conceded by the offense because of a sack by the player on this play (numeric)",
    "Whether or not the player forced a safety on this play (numeric)",
    "Whether or not the player recorded a solo tackle on this play (numeric)",
    "Whether or not the player was awarded an assisted tackle on this play (numeric)",
    "Whether or not the player recorded a tackle behind the line of scrimmage on this play (numeric)",
    "The yards conceded by the offense because of a tackle behind the line of scrimmage by the player on this play (numeric)",
    "Whether or not the player intercepted a pass on this play (numeric)",
    "The yards returned by the player on an intercepted pass on this play (numeric)",
    "The number of fumbles recovered by the player on this play (numeric)",
    "The yards returned by the player on a fumble recovery on this play (numeric)",
    "Whether or not the player was the initial pass rusher on this play (numeric)",
    "The names of all the penalties that were called on this player on this play (text)",
    "Boolean indicating whether the player pressured the QB, defined as achieving a peak pressure probability greater than or equal to 0.75 over the course of a dropback (Boolean)",
    "The time elapsed from snap to the first instance of this player reaching a pressure probability greater than or equal to 0.75 (numeric)",
    "The time it took for this player to cross the line of scrimmage as a pass rusher after the ball was snapped (numeric)",
    "Boolean indicating whether the player was in motion at snap (Boolean)",
    "Boolean indicating whether the player shifted since the lineset (Boolean)",
    "Boolean indicating whether the player went in motion after they were initially set at the line on this play (Boolean)",
    "Boolean indicating if the player was running a route on this play (Boolean)",
    "The name of the route ran by the player on this play (text)",
    "The NFL player ID of the primary opponent being blocked by the player on this play (numeric)",
    "The NFL player ID of the secondary opponent being blocked by the player on this play (numeric)",
    "The NFL player ID of the tertiary opponent being blocked by the player on this play (numeric)",
    "Whether or not any of the pass rushers that the blocker had a true matchup against recorded a pressure on this play (numeric)",
    "The time elapsed from snap to the first instance of a pass rusher who the blocker had a true matchup against achieving a pressure probability above 0.75 on this play (numeric)",
    "The specific defensive coverage assignment given to the player on this play (text)",
    "The NFL player ID of the opponent tagged as the primary matchup in coverage for the defender on this play (numeric)",
    "The NFL player ID of the opponent tagged as the secondary matchup in coverage for the defender on this play (numeric)"
  )
)

kable(
  player_play_dict,
  caption = "Table: Data Dictionary for `player_play.csv`"
)

```

# Missingness Visualizations and Data Cleaning

```{r}

player_play_missing <- player_play |>
  summarise(across(everything(), ~mean(is.na(.)) * 100)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "missing_percent") |>
  filter(missing_percent > 0) |>
  arrange(desc(missing_percent))

player_play_missing |>
  ggplot(aes(x = reorder(variable, -missing_percent), y = missing_percent)) +
  geom_bar(stat = "identity", fill = "#013369") +
  coord_flip() +
  labs(title = "Missingness in `player_play` Dataset",
       x = "Variable", y = "Percent Missing") +
  theme_minimal()
```

These missing pieces are largely related to what these variables are signifying. For instance the `timeToPresssureAsPassRusher` only applies to the offensive line positions that are blocking. Thus the main pattern of missingness here is position dependent, and with a data set featuring every position on the field for every play, its clear why there are many vars with high missing rates. 

```{r}

play_missing <- plays |>
  summarise(across(everything(), ~mean(is.na(.)) * 100)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "missing_percent") |>
  filter(missing_percent > 0) |>
  arrange(desc(missing_percent))

play_missing |>
  ggplot(aes(x = reorder(variable, -missing_percent), y = missing_percent)) +
  geom_bar(stat = "identity", fill = "#013369") +
  coord_flip() +
  labs(title = "Missingness in `plays` Dataset",
       x = "Variable", y = "Percent Missing") +
  theme_minimal()

```

The main pattern of missingness in this dataset comes from the type of play ran, when the play is a pass, all of the rush related variables are missing and vice versa. 

```{r}

players_missing <- players |>
  summarise(across(everything(), ~mean(is.na(.)) * 100)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "missing_percent") |>
  filter(missing_percent > 0) |>
  arrange(desc(missing_percent))

players_missing |>
  ggplot(aes(x = reorder(variable, -missing_percent), y = missing_percent)) +
  geom_bar(stat = "identity", fill = "#013369") +
  coord_flip() +
  labs(title = "Missingness in players Dataset",
       x = "Variable", y = "Percent Missing") +
  theme_minimal()
```

```{r}
games_missing <- games |>
  summarise(across(everything(), ~mean(is.na(.)) * 100)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "missing_percent") |>
  filter(missing_percent > 0) |>
  arrange(desc(missing_percent))

games_missing |>
  ggplot(aes(x = reorder(variable, -missing_percent), y = missing_percent)) +
  geom_bar(stat = "identity", fill = "#013369") +
  coord_flip() +
  labs(title = "Missingness in games Dataset",
       x = "Variable", y = "Percent Missing") +
  theme_minimal()
```

No missing vars in the `games` data set

Overall I'm not going to do anything in order to handle the missing data, because the missing observations are context dependent to their variable. 

## String Manipulation

### Expand Dropback Outcomes

```{r}
plays <- plays |>
  mutate(passResultFull = case_when(
    passResult == "C" ~ "Complete pass",
    passResult == "I" ~ "Incomplete pass",
    passResult == "S" ~ "Quarterback sack",
    passResult == "IN" ~ "Intercepted pass",
    passResult == "R" ~ "Scramble",
    TRUE ~ "Unknown"
  ))

plays|>
  group_by(passResultFull) |>
  summarise(Count = n(), .groups = "drop")
```

### Conference Labeling for Players

```{r}

add_conference <- function(college_name) {
  case_when(
    str_detect(college_name, "Alabama|Auburn|Georgia|Florida|LSU|Tennessee|Texas A&M|Vanderbilt|Arkansas|Mississippi|South Carolina|Oklahoma|Texas") ~ "SEC",
    str_detect(college_name, "Michigan|Ohio State|Penn State|Wisconsin|Iowa|Nebraska|Northwestern|Illinois|Indiana|Purdue|Minnesota|Rutgers|Maryland|Oregon|Washington|Southern California|UCLA") ~ "Big Ten",
    str_detect(college_name, "Clemson|Miami|Florida State|North Carolina|Virginia Tech|Pittsburgh|Duke|Georgia Tech|Louisville|NC State|Syracuse|Wake Forest|Stanford|California|Southern Methodist|Virginia") ~ "ACC",
    str_detect(college_name, "Baylor|Texas Christian|Texas Tech|Oklahoma State|Iowa State|West Virginia|Kansas|Kansas State|Cincinnati|Central Florida|Brigham Young|Houston|Colorado|Arizona|Arizona State|Utah") ~ "Big 12",
    TRUE ~ "Other"
  )
}

players <- players |>
  mutate(Conference = add_conference(collegeName))

conference_summary <- players |>
  group_by(Conference) |>
  summarize(PlayerCount = n(), .groups = "drop")

conference_summary

```


# Exploratory Data Analysis

## Tables of Summary Statistics and Merged Data

### Joined Data Analyzing RB Routes

```{r}
position_route_table <- player_play |>
  inner_join(players, by = "nflId") |>
  filter(position %in% c("RB", "TE", "WR"), !is.na(routeRan)) |>
  count(position, routeRan, name = "route_count") |>
  group_by(position) |>
  mutate(
    route_percentage = round((route_count / sum(route_count)) * 100, 2)
  ) |>
  arrange(position, desc(route_count)) |>
  ungroup()

position_route_table <- position_route_table |>
  select(position, routeRan, route_count, route_percentage)

position_route_table_flextable <- flextable(position_route_table)

position_route_table_flextable
```


### Analyzing the Type of Defenses Teams Run in AFC

```{r}
afc_teams <- c("BAL", "BUF", "CIN", "CLE", "DEN", "HOU", "IND", "JAX", 
               "KC", "LV", "LAC", "MIA", "NE", "NYJ", "PIT", "TEN")

afc_games <- games |>
  filter(homeTeamAbbr %in% afc_teams | visitorTeamAbbr %in% afc_teams)

afc_defensive_plays <- plays |>
  filter(defensiveTeam %in% afc_teams)

defense_summary <- afc_defensive_plays |>
  group_by(defensiveTeam, pff_passCoverage) |>
  summarise(Count = n(), .groups = "drop")

defense_percentage <- defense_summary |>
  group_by(defensiveTeam) |>
  mutate(Percent = round(Count / sum(Count) * 100, 2)) |>
  ungroup()

defense_wide <- defense_percentage |>
  select(defensiveTeam, pff_passCoverage, Percent) |>
  pivot_wider(names_from = pff_passCoverage, values_from = Percent, values_fill = 0)

defense_table <- defense_wide |>
  flextable() |>
  set_header_labels(defensiveTeam = "Team") |>
  set_caption("AFC Teams: Defensive Coverage Percentage Breakdown")|>
  theme_vanilla() |>
  autofit()

defense_table
  
```

### NFC Rush Yards Summary 

```{r}
nfc_teams <- c("ARI", "ATL", "CAR", "CHI", "DAL", "DET", "GB", "LAR", 
               "MIN", "NO", "NYG", "PHI", "SEA", "SF", "TB", "WAS")

nfc_rush_plays <- plays |>
  filter(possessionTeam %in% nfc_teams & !is.na(rushLocationType) 
         & is.na(penaltyYards))
rush_summary <- nfc_rush_plays |>
  group_by(possessionTeam, rushLocationType) |>
  summarize(AverageRushYards = round(mean(yardsGained, na.rm = TRUE), 2), .groups = "drop")

rush_summary_wide <- rush_summary |>
  pivot_wider(names_from = rushLocationType, values_from = AverageRushYards, values_fill = 0)

rush_table <- rush_summary_wide |>
  flextable() |>
  set_header_labels(possessionTeam = "Team") |>
  set_caption("NFC Teams: Average Rush Yards by Rush Location") |>
  theme_vanilla() |>
  autofit()

rush_table

```


## Data Visualizations

### Throwing Heat Map of Quarterbacks with more than 50 dropbacks

```{r}
filtered_plays <- plays |>
  filter(!is.na(targetX) & !is.na(targetY))

filtered_plays <- filtered_plays |>
  mutate(
    X_bin = cut(targetX, breaks = seq(0, 120, by = 1), include.lowest = TRUE, labels = FALSE), 
    Y_bin = cut(targetY, breaks = seq(0, 53.3, by = 1), include.lowest = TRUE, labels = FALSE)
  )

zone_counts <- filtered_plays |>
  group_by(X_bin, Y_bin) |>
  summarize(Count = n(), .groups = "drop")

zone_counts <- complete(zone_counts, X_bin = 1:120, Y_bin = 1:53, fill = list(Count = 0))

ggplot(zone_counts, aes(x = X_bin, y = Y_bin, fill = Count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "blue", high = "red") +
  geom_vline(xintercept = seq(0, 120, by = 10), color = "white", linetype = "dashed", size = 0.5) +
  geom_vline(xintercept = c(10, 110), color = "black", linetype = "dashed", size = 1) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "green4", color = "green4"),
    panel.grid = element_blank(),
    axis.text = element_text(color = "black", size = 8, face = "bold"),
    axis.title = element_text(color = "black", size = 10, face = "bold"),
    plot.title = element_text(color = "black", size = 12, hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.ticks = element_line(color = "white")
  ) +
  labs(
    title = "Heatmap of Passing Target Locations (All Quarterbacks)",
    x = "Field Length (Yards)",
    y = "Field Width (Yards)",
    fill = "Number of Throws"
  ) +
  coord_fixed()




```

Here we have a football field broken up into 1x1-yard bins that count the number of throws targeted in that area. Some takeaways from this show that between the 30-yard lines there are the most throws generally, which makes sense given that is where most of the game takes place. But speaking from a width perspective, it’s clear that along the hash marks to the sidelines are a much more prevalent target area for quarterbacks compared to the middle of the field. Another observation is that the throws into the end zone are hot around the corners, with the middle seeing far fewer targets overall.

PS I endlessly struggled to get the 50 yard line as the center of the chart but things seemed to fall apart in other places when scaling the x axis, and this was the best I could do.  

### EPA Per Play for AFC West by Week

```{r}

plays_with_week <- plays |>
  inner_join(games |> 
               select(gameId, week), by = "gameId")

afc_west_teams <- c("DEN", "KC", "LV", "LAC") 
afc_west_plays <- plays_with_week |>
  filter(possessionTeam %in% afc_west_teams & !is.na(expectedPointsAdded))

afc_west_epa <- afc_west_plays |>
  group_by(possessionTeam, week) |>
  summarize(AverageEPA = mean(expectedPointsAdded, na.rm = TRUE), .groups = "drop")

ggplot(afc_west_epa, aes(x = week, y = AverageEPA, color = possessionTeam, group = possessionTeam)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Average EPA Per Play by Game for AFC West Teams",
    x = "Week",
    y = "Average Expected Points Added (EPA)",
    color = "Team"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = 1:9) + 
  scale_color_manual(values = c("#FB4F14", "#E31837", "#0080C6", "#A5ACAF"))


```
The Best way to look at this graph is to compare it against the W/L record for each team through 9 weeks in 2022. The Broncos at that time had a 3-5 record, Raiders 2-6, Chargers 5-3, and Chiefs 6-2. By looking at the EPA per play on a week by week basis it would track that the Broncos and Raiders would be at the bottom, although somewhat surprisingly the Broncos were able to eek out one more win in the 9 weeks with their failure of an offense. The Chiefs having 0 weeks with negative EPA/Play is illustrated by the best record of the bunch. 


### Yards Gained by Coverage Type

```{r}
nfc_north_teams <- c("GB", "MIN", "CHI", "DET") 
nfc_north_plays <- plays |>
  filter(defensiveTeam %in% nfc_north_teams & !is.na(pff_passCoverage) & !is.na(yardsGained))

nfc_north_coverage <- nfc_north_plays |>
  group_by(defensiveTeam, pff_passCoverage) |>
  summarize(Snaps = n(), AverageYards = mean(yardsGained, na.rm = TRUE), .groups = "drop") |>
  filter(Snaps >= 30)

ggplot(nfc_north_coverage, aes(x = defensiveTeam, y = AverageYards, fill = pff_passCoverage)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_text(
    aes(label = Snaps), 
    position = position_dodge(width = 0.9), 
    vjust = -0.5,
    size = 3
  ) +
  labs(
    title = "Average Yards Gained by Coverage Type for NFC North (Min 30 Snaps)",
    x = "Team",
    y = "Average Yards Gained",
    fill = "Coverage Type"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```

There are a few interesting observations that come from this chart. First being that in the first half of 2022, Chicago and Green Bay played only three coverage for the majority of their snaps. Compare that to Minnesota who played 7 coverages over 30 times, with both cover 6 techniques leading to the worst results in the division on a per play basis. 



### Success Rate of the Denver Broncos by Quarter

```{r}
league_plays <- plays |>
  filter(!is.na(yardsToGo) & !is.na(yardsGained))

league_plays <- league_plays |>
  mutate(
    success = case_when(
      down == 1 & yardsGained >= 0.4 * yardsToGo ~ 1,
      down == 2 & yardsGained >= 0.6 * yardsToGo ~ 1,
      (down == 3 | down == 4) & yardsGained >= yardsToGo ~ 1,
      TRUE ~ 0
    )
  )

broncos_success <- league_plays |>
  filter(possessionTeam == "DEN" & quarter %in% 1:4) |>
  group_by(quarter, down) |>
  summarize(SuccessRate = mean(success, na.rm = TRUE) * 100, .groups = "drop") |>
  mutate(Type = "Broncos")

league_success <- league_plays |>
  filter(quarter %in% 1:4) |>
  group_by(quarter, down) |>
  summarize(SuccessRate = mean(success, na.rm = TRUE) * 100, .groups = "drop") |>
  mutate(Type = "League Average")

success_comparison <- bind_rows(broncos_success, league_success)

ggplot(success_comparison, aes(x = factor(quarter), y = SuccessRate, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~down, labeller = labeller(down = c("1" = "1st Down", "2" = "2nd Down", "3" = "3rd Down", "4" = "4th Down"))) +
  labs(
    title = "Success Rate by Down and Quarter: Broncos vs League Average",
    x = "Quarter",
    y = "Success Rate (%)",
    fill = "Team"
  ) +
  scale_fill_manual(values = c("Broncos" = "#FB4F14", "League Average" = "gray")) +
  theme_minimal()

```

Here is another visualization illustrating my beloved Bronco's anemic offense under Russell Wilson during the first half of 2022. Success rate is defined by pro-football-reference as "A play that gains at least 40% of yards required on 1st down, 60% of yards required on 2nd down, and 100% on 3rd or 4th down" We can see by this chart that the Broncos were below league average in every single instance in downs 1-3, and only higher in 4th down of the second quarter. This graph alone justifies eating the 50 million dollars in dead cap to get Russell Wilson off the team, no matter how successful he is operating the Steelers offense presently. 


### Time to Throw vs Yards Gained Interactive Graph

```{r}

nfc_teams <- c("DAL", "PHI", "NYG", "WAS", "GB", "CHI", "MIN", "DET", 
               "TB", "NO", "CAR", "ATL", "SF", "SEA", "ARI", "LAR")

nfc_completed_passes <- plays |>
  filter(
    possessionTeam %in% nfc_teams & 
    passResult == "C" & 
    !is.na(timeToThrow) & 
    !is.na(yardsGained)
  ) |>
  mutate(
    playDescriptionShort = substr(playDescription, 1, 50)
  )

correlation <- cor(nfc_completed_passes$timeToThrow, nfc_completed_passes$yardsGained, use = "complete.obs")

scatter_plot <- 
  ggplot(nfc_completed_passes, aes(x = timeToThrow, y = yardsGained, color = possessionTeam)) +
  geom_point(aes(text = paste(
    "Team:", possessionTeam,
    "<br>Yards Gained:", yardsGained,
    "<br>Time to Throw:", timeToThrow,
    "<br>Play Description:", playDescriptionShort
  )), alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", color = "red", linetype = "dashed", size = 1) +
  annotate("text", x = max(nfc_completed_passes$timeToThrow) * 0.7, 
           y = max(nfc_completed_passes$yardsGained) * 0.9, 
           label = paste("Correlation: ", round(correlation, 2)), 
           size = 4, color = "black") +
  labs(
    title = "NFC Teams: Time to Throw vs. Yards Gained (Completed Passes)",
    x = "Time to Throw (Seconds)",
    y = "Yards Gained",
    color = "Team"
  ) +
  theme_minimal()

interactive_plot <- ggplotly(scatter_plot, tooltip = "text")
interactive_plot


```
Somewhat to my surprise, the correlation between time to throw and yards gained from completed passes is pretty low at 0.29, I imagine if I included the incomplete passes in the regression, the r value would approach 0. Perhaps this is due to the fact that, while QBs may have more time, the primary job of the play is to hit your first read and accomplish the task of the play design, as opposed to more time allowing for players to get down field more. I ran this without shortening the play description and my computer almost exploded, so the limit to 50 characters is due to that reason. 

#Monte Carlo Methods of Inference

```{r}
# Filter data for dropback plays with valid yards gained and play action labels
play_action_data <- plays |>
  filter(!is.na(playAction) & isDropback == TRUE & !is.na(yardsGained)) |>
  mutate(playAction = ifelse(playAction, "Play Action", "No Play Action"))

# Calculate observed test statistic (difference in medians)
observed_diff <- play_action_data |>
  group_by(playAction) |>
  summarize(median_yards = median(yardsGained)) |>
  summarize(diff = diff(median_yards)) |>
  pull(diff)

# Perform permutation test
set.seed(1999)
n_permutations <- 1000
null_distribution <- replicate(n_permutations, {
  shuffled_data <- play_action_data |>
    mutate(playAction = sample(playAction))
  shuffled_diff <- shuffled_data |>
    group_by(playAction) |>
    summarize(median_yards = median(yardsGained)) |>
    summarize(diff = diff(median_yards)) |>
    pull(diff)
  return(shuffled_diff)
})

# Calculate p-value
p_value <- mean(abs(null_distribution) >= abs(observed_diff))

# Summarize medians for each group
group_medians <- play_action_data |>
  group_by(playAction) |>
  summarize(median_yards = median(yardsGained), n = n())

# Extract medians for Play Action and No Play Action
play_action_median <- group_medians |>
  filter(playAction == "Play Action") |>
  pull(median_yards)
no_play_action_median <- group_medians |>
  filter(playAction == "No Play Action") |>
  pull(median_yards)

# Print results
cat("Play Action Median Yards Gained:", round(play_action_median, 2), "\n")
cat("No Play Action Median Yards Gained:", round(no_play_action_median, 2), "\n")
cat("Observed Difference in Median Yards Gained:", round(observed_diff, 2), "\n")

# Visualize null distribution
library(ggplot2)
ggplot(data = data.frame(null_distribution), aes(x = null_distribution)) +
  geom_histogram(binwidth = 0.5, fill = "lightblue", color = "black") +
  geom_vline(xintercept = observed_diff, color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = quantile(null_distribution, probs = c(0.025, 0.975)), 
             color = "blue", linetype = "dotted", size = 1) +
  labs(
    title = "Null Distribution of Median Yards Gained Difference: Play Action vs No Play Action",
    x = "Difference in Median Yards Gained",
    y = "Frequency"
  ) +
  theme_minimal()

# Visualize observed distributions with density plots
ggplot(play_action_data, aes(x = yardsGained, fill = playAction)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("Play Action" = "blue", "No Play Action" = "red")) +
  geom_vline(aes(xintercept = median_yards), 
             data = group_medians, 
             color = c("blue", "red"), 
             linetype = "dashed", size = 1) +
  coord_cartesian(xlim = c(-10, 20)) +
  labs(
    title = "Distribution of Yards Gained: Play Action vs No Play Action",
    x = "Yards Gained",
    y = "Density",
    fill = "Play Type"
  ) +
  theme_minimal(base_size = 14)

# Interpret results
if (p_value < 0.05) {
  print(paste("The observed difference in median yards gained is statistically significant (p =", round(p_value, 3), ")."))
} else {
  print(paste("The observed difference in median yards gained is not statistically significant (p =", round(p_value, 3), ")."))
}

```


# Conclusions / Main Takeaways

I found this exercise incredibly rewarding overall. I initially intended to have my data analyses follow a specific trend and theme, but I found that the 9 week sample was not enough to dive into any particular areas with any certainty. So instead I just went with my feel and curiosity in exploring different areas of the game that I found interesting. I think I created interesting visualizations that looked at all facets of the game of football, as well as diving into team and division analyses. Appreciate the great semester and I hope we get to connect in the future in other classes or research oppurtunities. 




